Inferring generation-interval distributions from contact tracing data: \\ \emph{in prep, PRSB}

### 2019-09-23 doc/response.tex

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}

\usepackage{color}

\usepackage{lmodern}
\usepackage{amssymb,amsmath}

\newcommand{\rR}{\mbox{$r$--$\cal R$}}
\newcommand{\RR}{\ensuremath{{\cal R}}}
\newcommand{\RRhat}{\ensuremath{{\hat \cal R}}}
\newcommand{\Rx}[1]{\ensuremath{{\cal R}_{#1}}} 
\newcommand{\Ro}{\Rx{0}}
\newcommand{\Reff}{\Rx{\mathit{eff}}}
\newcommand{\Tc}{\ensuremath{C}}

\newcommand{\rev}{\subsection*}
\newcommand{\revtext}{\textsf}
\setlength{\parskip}{\baselineskip}
\setlength{\parindent}{0em}

\newcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textsl{#3}\textbf{]}}}
\newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}
\newcommand{\dc}[1]{\comment{blue}{DC}{#1}}
\newcommand{\jsw}[1]{\comment{green}{JSW}{#1}}
\newcommand{\hotcomment}[1]{\comment{red}{HOT}{#1}}

\begin{document}

\noindent Dear Editor:

Thank you for the chance to transfer this submission. 
Below please find our responses from the previous round.

\rev{Reviewer 1}

\revtext{This paper considers estimation of generation interval distributions from data, paying particular attention to the questions of censoring. These are important questions, and the paper clearly lays out the theory of generation times and its implications.}

\revtext{My main concern about the paper is audience; for a typical Proceedings B / applied audience, there is no work with real epidemiological data (although there is real network data); but for a more theoretical audience, the new analytic results are not quite as general as would be expected.}

\revtext{Given that the questions considered are important and worthwhile, my suggestion is that the paper should be revised with a substantial additional piece of work added involving work with a real epidemic dataset (not necessarily novel data) to demonstrate the theoretical developments, with R code in a tidier / more user-friendly format than is currently on the associated GitHub site.
}

As per editor's suggestion, we have transferred our submission to Interface journal, which better suits our manuscript in its current form. Due to time constraints, we will not be able to make substantial changes to our manuscript.

Generation-interval-based approaches (Euler-Lotak equation) are still widely used in outbreak analyses to estimate the reproduction number. While some of the analytical results that we present here (e.g., egocentric kernel) may not be as general as would be expected, we expect biological intuitions (the effects of censoring and spatial reduction on generation intervals and reproduction number) that we get from our results to be fairly general.

\revtext{Some minor comments:}

\revtext{Introduction. “Can be characterized by”. These are properties but don’t uniquely determine the epidemic.}

We have followed the reviewer's comment in changing the text:

``An epidemic can be described by the exponential growth rate, $r$, and the reproductive number, \RR.''

\revtext{Fig 1 -> two things change between each figure. I would describe the left hand as an individual realisation of the random process, and the right hand as a population-level behaviour emerging from the random process.}

We have followed the reviewer's comment in changing the figure caption:

``(Left) an individual-level kernel of an infected individual with latent period of 11.4 days followed by infectious period of 5 days;
this represents an individual realization of a random process.
(Right) a population-level kernel of infected individuals with latent and infectious periods exponentially distributed with means of 11.4 and 5 days, respectively;
this represents a population average of a random process.''

\revtext{Eq(15) -> typo on last line of the equations.}

Fixed.

\rev{Reviewer 2}

\revtext{The manuscript addresses very important questions. Some comments:}

\revtext{The same notation, e.g. (3) is used for equation three and reference 3 (and also some effects mention in beginning of Sec 5). Use different notation.}

Fixed.

\revtext{p5, eq 3: perhaps $\RR \rightarrow \RR(t)$?}

We made minor changes to this equation for clarity:

\begin{equation*}
i(t) = S(t) \int K(s) i(t-s) ds = \RR_0 S(t) \int g(s) i(t-s) ds.
\end{equation*}

\revtext{p5, eq 5: I think S(t) should be removed as it is contained in R?}

This was a typo. We changed $\RR$ to $\RR_0$:

\begin{equation*}
i_{t-\tau}(t) = \RR_0 i(t-\tau) g(\tau) S(t)
\end{equation*}

\revtext{p5-6: perhaps mention that $t, s$ is for calendar time and $\tau$ time relative to/from infection.}

We added the following sentence in the beginning of section 2 to clarify this:

``Hereafter, we use $t, s$ to represent calendar time and $\tau$ to represent time relative to/from infection.''

\revtext{p6, top: when you say and write "proportional" there are 3 time parameters. Be explicit and write: with respect to $\tau$}

Fixed.

\revtext{p8, top and later: I have a problem with $N(a)$. Does your model allow for different number of acquaintances? But then your $\RR_0$ is no longer valid. Also, you are always assuming neighbours/acquaintances are chosen at random - please stress this more.}

We added the following sentences in the beginning of section 3 to make this clear:

``To explore the effect of multiple contacts on realized generation intervals, we no longer assume that the population is homogeneous (and therefore do not rely on the classical definition of the basic reproductive number).
Instead, we assume that a disease spreads on a network;
infected individuals contact their ``acquaintances'' at random, but ``acquantanceships'' are predetermined by the network structure before the beginning of an epidemic.''

\revtext{p8, bottom: I did not follow how you got $\hat g$. What is a, $N(a)$ and $f(a)$? Explain more.}

We tried to clarify this.

``... number of acquaintances $N(a)$ ...''

``The population-level egocentric kernel is found by integrating the individual-level egocentric over individual variations:
\begin{equation}\label{eq:ego}
\hat{K}(\tau) = \int \hat{k}(\tau; a) f(a) da,
\end{equation}
where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
Essentially, the population-level egocentric kernel further accounts for the probability that a susceptible individual has not been infected within an infected-susceptible pair.
Trapman \textit{et al.} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
The population-level egocentric generation-interval distribution is:
\begin{equation}
\hat{g}(\tau) = \frac{\hat{K}(\tau)}{\int \hat{K}(s) ds}.
\label{eq:conditional}
\end{equation}
The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized from an \emph{average} infected-susceptible pair ...''

\revtext{p9, bottom: what is a homogeneous network? Complete network, alla have same degree or ...? Explain better why your networks of size 5 matter in a large community (local aspects?)}

We tried to clarify this.

``
Simulations on a small homogeneous network (i.e., complete network) confirm this additional effect (Fig. 3, right panel). We can think of simulations on this network as an approximation of (local) infection process in a small household, consisting of 5 individuals;
we expect realistic local network structures (and their effects on the realized generation intervals) to lay between a star network and a complete network...''

\revtext{Sec 5: some of the 4 effects described also affect R whereas others don't. Be more clear to the reader. For instance, the ordinare Euler-Lotka equations is not fulfilled in network situation. I think Ball, Pellis and Trapman have investigated different reproduction numbers for network epidemics and related.}

All 4 effects affect $\RR$. Our simulations show that Euler-Lotka equation can work even in network situation, as long as we can find the proper generation-interval distribution (by using temporal correction early in an epidemic). We tried to make this point clear.

``When generation intervals are sampled through contact tracing, there will be four effects present in the sample: (1) right-censoring effect, (2) egocentric depletion effect, (3) local depletion effect and (4) global depletion effect.
We can correct explicitly for the egocentric effect and, in the case of exponential growth, the right-censoring effect;
we showed that these effects shorten the mean observed generation intervals, which in turn will reduce the estimate of the reproduction number.
While the other two effects are difficult to measure, we can make qualitative predictions about their effects on the realized generation intervals and reproductive numbers: 
both local and global depletion effects also reduce number of infections that occur and shorten generation intervals.''

``When temporal correction is performed early in an outbreak, during the exponential growth phase, the effective distribution should incorporate egocentric and local spatial effects but not the global effects; we expect this distribution to correctly link $r$ and $\RR$ through the Euler-Lotka equation.''

\revtext{In your analysis you assume that you observe the sampled contact traced generation times without error. It is known (e.g. your ref 18) that onset of symptoms rather than infection time are observed, and that often there are several potential infectors. You don't have to include this in your analysis, but at least it should be acknowledged.}

We added the following sentences:

``For brevity, we assume that \emph{all} generation intervals (no under-reporting) are observed without error.
In practice, it is difficult to measure generation intervals precisely because (1) infection events are often unobserved and (2) there may be multiple potential infectors; these factors can introduce biases to the estiamtes of the reproduction number.
We do not pursue these directions in this study.''

\revtext{p14, sec 7.2: the $S_i$ should vary over time. When is it measured in the algorithm?}

We added the following sentence:

``... (and therefore decreasing $S_i$ by 1 for all individuals $i$ that are acquaintances of the newly infected individual) ...''

\revtext{p17, middle. Suddenly you talk about under-reporting. This is another very important area which would affect your analysis in many ways. Mention that you neglect this elsewhere.
}

We added the following sentence:

``Here, we assume that \emph{all} generation intervals (no under-reporting) are observed without error''

\end{document}

### 2019-09-23 doc/network.tex
stuff

In a homogeneously mixing population, current disease incidence at time $t$, $i(t)$, is the product of the current infectiousness of individuals infected in the past and the current proportion of the population susceptible, $S(t)$.
\begin{equation}
i(t) = S(t) \int K(s) i(t-s) ds = \RR(t) \int g(s) i(t-s) ds,
\end{equation}
where $\RR = S(t) \RR_0$ is the effective reproductive number.
This model, referred to as the renewal equation, can describe a wide range of epidemic models \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, wallinga2007generation, roberts2007model}.
Over a period of time where the susceptible proportion $S$ remains roughly constant, we would expect approximately exponential growth in incidence $i(t)$; assuming $i(t) = i(0) \exp(r t)$ yields the Euler-Lotka equation \citep{lotka1907relation}, which provides a direct link between speed and strength of an epidemic:
\begin{equation}
\frac{1}{\RR} = \int g(s) \exp(-r s) ds.
\end{equation}

@@

\rev{I did not follow how you got $\hat g$. What is a, $N(a)$ and $f(a)$? Explain more.}
\swp{Reviewer 2 was not happy with this description. Come back.}
The population-level egocentric kernel is found by integrating over individual variations:
\begin{equation}\label{eq:ego}
\hat{K}(\tau) = \int \hat{k}(\tau; a) f(a) da,
\end{equation}
where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
Trapman \textit{et al.} \cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
The population-level egocentric generation-interval distribution is:
\begin{equation}
\hat{g}(\tau) = \frac{\hat{K}(\tau)}{\int \hat{K}(s) ds}.
\label{eq:conditional}
\end{equation}
The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized by an \emph{average} primary case; for convenience, we will often omit ``population-level''.
Finally, the link between the growth rate and the egocentric reproductive number are linked by the egocentric generation-interval distribution (and the Euler-Lotka equation) \citep{trapman2016inferring}:
\begin{equation}
\frac{1}{\hat{\RR}} = \int \hat{g}(\tau) \exp(-r \tau) d\tau.
\label{eq:egorR}
\end{equation}
As the egocentric distribution always has a shorter mean than the intrinsic distribution, \Rhat\ will always be smaller than $\RR$ estimated from the intrinsic distribution;
this generation-interval-based argument provides a clear biological interpretation for the result presented by \cite{trapman2016inferring}.
% Similarly, \Rhat\ will be greater than the \emph{true} reproductive number \RR\, since it does not account for depletion of susceptibles by other routes.

@@

\section{Inferring generation-interval distributions from contact-tracing data}
\rev{
some of the 4 effects described also affect R whereas others don't. Be more clear to the reader. For instance, the ordinare Euler-Lotka equations is not fulfilled in network situation. I think Ball, Pellis and Trapman have investigated different reproduction numbers for network epidemics and related.
}
\swp{They all affect R. I guess we weren't clear enough. Also, the point of our paper is that we can use 
Euler-Lotka equations as long as we know which distribution to use. I tried to clarify these points.}

When generation intervals are sampled through contact tracing, there will be four effects present in the sample: (1) right-censoring effect, (2) egocentric depletion effect, (3) local depletion effect and (4) global depletion effect.
We can correct explicitly for the egocentric effect and, in the case of exponential growth, the right-censoring effect;
\swp{we showed that these effects shorten the mean observed generation intervals, which in turn will reduce the estimate of the reproduction number.}
While the other two effects are difficult to measure, we can make qualitative predictions about their effects on the realized generation intervals and reproductive numbers: 
both local and global depletion effects also reduce number of infections that occur and shorten generation intervals.

@@

\begin{figure}[!pbth]
\includegraphics[width=\textwidth]{../fig/cmp_reproductive.pdf}
\caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
Using the observed generation-interval distributions (based on the first 1000 generation intervals) without correcting for right-censoring severely underestimates the reproductive number.
Similarly, using the intrinsic generation-interval distribution overestimates the reproductive number because it fails to account for local spatial effects; the egocentric distribution corrects for this only partially.
Both population-level and individual-level methods provide estimates of reproductive number that are consistent with the empirical estimates, which we define as the average number of secondary cases generated by the first 100 infected individuals, but the individual-level method is more precise.
\swp{Here, we assume that \emph{all} generation intervals (no under-reporting) are observed without error.}
Boxplots are generated using 100 stochastic simulations of the SEIR model on an empirical network using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 5 \textrm{ days}$ and mean infectious period $1/\gamma = 11.4 \textrm{ days}$. 
Per-pair contact rate $\lambda = 0.08 \textrm{ days}^{-1}$ is chosen to be sufficiently high such that differences are clear.
}
\label{fig:cmp}
\end{figure}

@@

To test these methods, we simulate 100 epidemics with Ebola-like parameters on an empirical network \citep{leskovec2016snap}
and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
As expected, calculating reproductive number based on the intrinsic generation-interval distribution overestimates the empirical reproductive number;
estimates based on the egocentric generation-interval distribution \eref{conditional} address this problem only partially, as they do not account for indirect (local) spatial effects. 
Direct estimates based on the observed generation intervals from contact tracing severely underestimate the empirical estimates.
While both population- and individual-level corrections provide similar estimates to the empirical reproductive number on average,
population-level estimates are more variable as they are more sensitive to outliers in generation intervals and our estimate of the growth rate.
For smaller values of \RR, we expect the differences to become smaller.
In the electronic supplementary material, we present the same figure using smaller \RR\ (Appendix A.1) and using Erlang-distributed latent periods (Appendix A.2), which better corresponds to Ebola.
Overall, our qualitative conclusions do not change.

@@

After repeating the contact process for all initially infected individuals, all contacts are put into a sorted queue.
The first person in the queue becomes infected (and therefore $S_i$ is decreased by 1), and the current time is updated to infection time of this individual. 
Any subsequent contacts made to this individual are removed from the queue because they will no longer be effective.
We repeat the contact process for this newly infected individual.
Then, new contacts are added to the sorted queue.
The simulation continues until there are no more contacts are left in the queue.

@@

The \textbf{Population-level method} estimates the effective generation-interval distribution by reversing the inverse exponential weighting in the observed generation-interval distribution:
\begin{equation}
\tsub{g}{eff}(\tau) \propto c_t(\tau) \exp(r\tau).
\end{equation}
In order to do so, we first fit a gamma distribution to the observed generation-interval distributions; specifically, we estimate the mean $\bar G$ and shape $\alpha$ of a gamma distribution by maximum likelihood.
Then, the effective generation-interval distribution also follows a gamma distribution with mean $\alpha/(\alpha/\bar G - r)$ and shape $\alpha$.

The \textbf{Individual-level method} models each infection $i$ from an infected individual $j$ as a non-homogeneous Poisson process between the time at which infector $j$ was infected ($t_j$) and the censorship time ($\tsub{t}{censor}$), with time-varying Poisson rate at time $t$ is equal to $\RR \tsub{g}{eff}(t - t_j)$, where $\tsub{g}{eff}(t)$ is the effective generation-interval distribution \citep{daley2007introduction}.
We use a gamma distribution (parameterized by its mean and shape) to model the effective generation-interval distribution.
Then, the probability that an individual $j$ infects $n_j$ individuals between $t_j$ and $\tsub{t}{censor}$ is equal to
\begin{equation}
\frac{\RR^{n_j} \exp \big(- \RR\, \tsub{G}{eff}(\tsub{t}{censor} - t_j ;\theta) \big) \prod_{i=1}^{n_j} \tsub{g}{eff}(s_{i, j}; \theta)}{n_j!},
\end{equation}
where $s_{i,j}$ is the observed generation interval between infector $j$ and infectee $i$, and $\theta$ is a (vector) parameter of the generation-interval distribution $g$ (and the corresponding cumulative distribution function $\tsub{G}{eff}$).

>>>>

In a homogeneously mixing population, current disease incidence at time $t$, $i(t)$, is the product of the current infectiousness of individuals infected in the past and the current proportion of the population susceptible, $S(t)$.
\begin{equation}
i(t) = S(t) \int K(s) i(t-s) ds = \RR_0 S(t) \int g(s) i(t-s) ds.
\end{equation}
This model, referred to as the renewal equation, can describe a wide range of epidemic models \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, wallinga2007generation, roberts2007model}.
Over a period of time where the susceptible proportion $S(t) = S$ remains roughly constant, we would expect approximately exponential growth in incidence $i(t)$; assuming $i(t) = i(0) \exp(r t)$ yields the Euler-Lotka equation \citep{lotka1907relation}, which provides a direct link between speed and strength of an epidemic:
\begin{equation}
\frac{1}{\RR} = \int g(s) \exp(-r s) ds,
\end{equation}
where $\RR = S(t) \RR_0$ is the effective reproductive number.

@@

\rev{I did not follow how you got $\hat g$. What is a, $N(a)$ and $f(a)$? Explain more.}
The population-level egocentric kernel is found by integrating the individual-level egocentric over individual variations:
\begin{equation}\label{eq:ego}
\hat{K}(\tau) = \int \hat{k}(\tau; a) f(a) da,
\end{equation}
where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
\swp{Essentially, the population-level egocentric kernel further accounts for the probability that a susceptible individual has not been infected within an infected-susceptible pair.}.
Trapman \textit{et al.} \cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
The population-level egocentric generation-interval distribution is:
\begin{equation}
\hat{g}(\tau) = \frac{\hat{K}(\tau)}{\int \hat{K}(s) ds}.
\label{eq:conditional}
\end{equation}
The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized from an \emph{average} infected-susceptible pair; for convenience, we will often omit ``population-level''.
Finally, the link between the growth rate and the egocentric reproductive number are linked by the egocentric generation-interval distribution (and the Euler-Lotka equation) \citep{trapman2016inferring}:
\begin{equation}
\frac{1}{\hat{\RR}} = \int \hat{g}(\tau) \exp(-r \tau) d\tau.
\label{eq:egorR}
\end{equation}
As the egocentric distribution always has a shorter mean than the intrinsic distribution, \Rhat\ will always be smaller than $\RR$ estimated from the intrinsic distribution;
this generation-interval-based argument provides a clear biological interpretation for the result presented by \cite{trapman2016inferring}.
% Similarly, \Rhat\ will be greater than the \emph{true} reproductive number \RR\, since it does not account for depletion of susceptibles by other routes.

@@

\section{Inferring generation-interval distributions from contact-tracing data}
\rev{
some of the 4 effects described also affect R whereas others don't. Be more clear to the reader. For instance, the ordinare Euler-Lotka equations is not fulfilled in network situation. I think Ball, Pellis and Trapman have investigated different reproduction numbers for network epidemics and related.
}

When generation intervals are sampled through contact tracing, there will be four effects present in the sample: (1) right-censoring effect, (2) egocentric depletion effect, (3) local depletion effect and (4) global depletion effect.
We can correct explicitly for the egocentric effect and, in the case of exponential growth, the right-censoring effect;
\swp{we showed that these effects shorten the mean observed generation intervals, which in turn will reduce the estimate of the reproduction number \cite{wallinga2007generation, park2019practical}.}
While the other two effects are difficult to measure, we can make qualitative predictions about their effects on the realized generation intervals and reproductive numbers: 
both local and global depletion effects also reduce number of infections that occur and shorten generation intervals.

@@

\begin{figure}[!pbth]
\includegraphics[width=\textwidth]{../fig/cmp_reproductive.pdf}
\caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
Using the observed generation-interval distributions (based on the first 1000 generation intervals) without correcting for right-censoring severely underestimates the reproductive number.
Similarly, using the intrinsic generation-interval distribution overestimates the reproductive number because it fails to account for local spatial effects; the egocentric distribution corrects for this only partially.
Both population-level and individual-level methods provide estimates of reproductive number that are consistent with the empirical estimates, which we define as the average number of secondary cases generated by the first 100 infected individuals, but the individual-level method is more precise.
Boxplots are generated using 100 stochastic simulations of the SEIR model on an empirical network using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 5 \textrm{ days}$ and mean infectious period $1/\gamma = 11.4 \textrm{ days}$. 
Per-pair contact rate $\lambda = 0.08 \textrm{ days}^{-1}$ is chosen to be sufficiently high such that differences are clear.
}
\label{fig:cmp}
\end{figure}

@@

To test these methods, we simulate 100 epidemics with Ebola-like parameters on an empirical network \citep{leskovec2016snap}
and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
\swp{
For brevity, we assume that \emph{all} generation intervals (no under-reporting) are observed without error.
In practice, it is difficult to measure generation intervals precisely because (1) infection events are often unobserved and (2) there may be multiple potential infectors; these factors can introduce biases to the estiamtes of the reproduction number \cite{britton2019estimation}.
We do not pursue these direction in this study.
}

As expected, calculating reproductive number based on the intrinsic generation-interval distribution overestimates the empirical reproductive number;
estimates based on the egocentric generation-interval distribution \eref{conditional} address this problem only partially, as they do not account for indirect (local) spatial effects. 
Direct estimates based on the observed generation intervals from contact tracing severely underestimate the empirical estimates.
While both population- and individual-level corrections provide similar estimates to the empirical reproductive number on average,
population-level estimates are more variable as they are more sensitive to outliers in generation intervals and our estimate of the growth rate.
For smaller values of \RR, we expect the differences to become smaller.
In the electronic supplementary material, we present the same figure using smaller \RR\ (Appendix A.1) and using Erlang-distributed latent periods (Appendix A.2), which better corresponds to Ebola.
Overall, our qualitative conclusions do not change.

@@

After repeating the contact process for all initially infected individuals, all contacts are put into a sorted queue.
The first person in the queue becomes infected \swp{(and therefore decreasing $S_i$ by 1 for all individuals $i$ that are acquaintances of the newly infected individual)}, and the current time is updated to infection time of this individual. 
Any subsequent contacts made to this individual are removed from the queue because they will no longer be effective.
We repeat the contact process for this newly infected individual.
Then, new contacts are added to the sorted queue.
The simulation continues until there are no more contacts are left in the queue.

@@

The \textbf{population-level method} estimates the effective generation-interval distribution by reversing the inverse exponential weighting in the observed generation-interval distribution:
\begin{equation}
\tsub{g}{eff}(\tau) \propto c_t(\tau) \exp(r\tau).
\end{equation}
In order to do so, we first fit a gamma distribution to the observed generation-interval distributions; specifically, we estimate the mean $\bar G$ and shape $\alpha$ of a gamma distribution by maximum likelihood.
Then, the effective generation-interval distribution also follows a gamma distribution with mean $\alpha/(\alpha/\bar G - r)$ and shape $\alpha$.

The \textbf{individual-level method} models each infection $i$ from an infected individual $j$ as a non-homogeneous Poisson process between the time at which infector $j$ was infected ($t_j$) and the censorship time ($\tsub{t}{censor}$), with time-varying Poisson rate at time $t$ is equal to $\RR \tsub{g}{eff}(t - t_j)$, where $\tsub{g}{eff}(t)$ is the effective generation-interval distribution \citep{daley2007introduction}.
We use a gamma distribution (parameterized by its mean and shape) to model the effective generation-interval distribution.
Then, the probability that an individual $j$ infects $n_j$ individuals between $t_j$ and $\tsub{t}{censor}$ is equal to
\begin{equation}
\frac{\RR^{n_j} \exp \big(- \RR\, \tsub{G}{eff}(\tsub{t}{censor} - t_j ;\theta) \big) \prod_{i=1}^{n_j} \tsub{g}{eff}(s_{i, j}; \theta)}{n_j!},
\end{equation}
where $s_{i,j}$ is the observed generation interval between infector $j$ and infectee $i$, and $\theta$ is a (vector) parameter of the generation-interval distribution $g$ (and the corresponding cumulative distribution function $\tsub{G}{eff}$).
